---
title: "NYPD Shooting"
author: "Sean Baier"
date: "`r Sys.Date()`"
output: 
    pdf_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r load_tidyverse, echo=TRUE, include=FALSE}
library(tidyverse)
library(lubridate)
library(modelr)
```

# Assignment

Import, tidy and analyze the NYPD Shooting Incident data set obtained. 

- Be sure your project is reproducible and contains some visualization and analysis. 
- You may use the data to do any analysis that is of interest to you. 
- You should include at least two visualizations and one model. 
- Be sure to identify any bias possible in the data and in your analysis.


## Load dependencies


```
library(tidyverse)
library(lubridate)
library(modelr)
```


## Raw Data for NYPD shootings

Import raw data from url.

<https://data.cityofnewyork.us/api/views/833y-fsy8/rows.csv?accessType=DOWNLOAD>

```{r get_nypd_data, echo=TRUE}
shootings_data <- read_csv("https://data.cityofnewyork.us/api/views/833y-fsy8/rows.csv?accessType=DOWNLOAD")
```


## Tidy Raw Shootings Data

- Some of the `perp_race` values are implicitly `UNKNOWN` due to them having the value `NA`. We are going to make these values explicitly `UNKNOWN` by mutating them.
- Normalize column names to lowercase and snake_case format.
- Remove unnecessary coordinate related columns since they are not needed for the following analysis.


```{r, tidy_shootings_data, echo=TRUE}
tidy_data <- shootings_data %>% 
  rename(
    full_date = OCCUR_DATE, 
    time = OCCUR_TIME, 
    neighborhood = BORO, 
    precinct = PRECINCT, 
    jurisdiction_code = JURISDICTION_CODE,
    statistical_murder = STATISTICAL_MURDER_FLAG,
    perp_age = PERP_AGE_GROUP,
    perp_sex = PERP_SEX,
    perp_race = PERP_RACE,
    vic_age = VIC_AGE_GROUP,
    vic_sex = VIC_SEX,
    vic_race = VIC_RACE) %>% 
  mutate(date = mdy(full_date)) %>%
  separate(date, into =c("year", "month", "day")) %>%
  replace_na(list(perp_race = "UNKNOWN")) %>%
  select(-c(X_COORD_CD, Y_COORD_CD, Latitude, Longitude, Lon_Lat, INCIDENT_KEY, LOCATION_DESC))
tidy_data
```



## Total number of shootings per month by neighborhood

We are also going to separate the `date` into `month`, `day`, and `year` in order to analyze the per month shootings per neighborhood.

```{r, total_shootings_per_month, echo=TRUE}
plot <- ggplot(data = tidy_data) +
  geom_bar(mapping = aes(x = month, fill = neighborhood), position = "dodge")

plot + theme(
  axis.text = element_text(size = rel(0.5))
)
```

## Shootings by race in neighborhoods

- Some potential bias might be to assume certain races might have higher numbers due to the fact that there is a higher population of those races in the neighborhood.
- It might also be assumed that the reason for less complete data on the `perp_race` is because they may have escaped arrest.

### Shootings by victim's race in neighborhoods

```{r, shootings_by_vic_race, echo=TRUE}
plot <- ggplot(data = tidy_data) + 
  geom_bar(mapping = aes(x = neighborhood, fill = vic_race), position = "dodge")

plot + theme(
  axis.text = element_text(size = rel(0.5))
)
```

### Shootings by perpetrator race in neighborhoods

```{r, race_neighborhoods, echo=TRUE}
plot <- ggplot(data = tidy_data) + 
  geom_bar(mapping = aes(x = neighborhood, fill = perp_race), position = "dodge")

plot + theme(
  axis.text = element_text(size = rel(0.5))
)
```

## Compare shootings to deaths per neighborhood

```{r, shootings_to_deaths, echo=TRUE}
shootings_to_deaths <- tidy_data %>%
  group_by(neighborhood, month) %>%
  mutate(shootings = n(), deaths = sum(statistical_murder)) %>%
  select(neighborhood, shootings, statistical_murder, deaths, month, full_date) %>%
  ungroup() %>%
  summarize(neighborhood, month, shootings, statistical_murder, deaths, full_date)
```

```{r, plot_shooting_to_deaths, echo=TRUE}
plot <- ggplot(data = shootings_to_deaths) + 
  geom_bar(mapping = aes(x = neighborhood, fill = statistical_murder), position = "dodge")

plot + theme(
  axis.text = element_text(size = rel(0.5))
)
```


# Model shootings to deaths


```{r, neighborhood_model, echo=TRUE}
data <- tidy_data %>%
  group_by(neighborhood, month) %>%
  mutate(neighborhood, month, shootings = n(), deaths = sum(statistical_murder)) %>%
  ungroup()

mod = lm(deaths ~ shootings, data = data)
summary(mod)
```


```{r, plot_model, echo=TRUE}
data %>% 
  mutate(pred = predict(mod)) %>%
  ggplot() +
  geom_point(aes(x = shootings, y = deaths), color = "blue") +
  geom_point(aes(x = shootings, y = pred), color = "red")
```
